Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2024-09-05 Manuscript/endo_herbarium.tex
removing comments for submitting to BioRxiv

	\textit{Manuscript elements}: Figure~1, figure~2, table~1, appendix~A (for print; including figure~A1, figure~A2, and table~A1), supplemental PDF. Figure~2 is to print in color.

@@

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. \ref{fig:map}, Fig. \ref{fig:temporal}A, Fig. \ref{fig:endo_status_map}).
%Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
On average we scored $4.7$ seeds per specimen of \emph{A. hyemalis}, $4.2$ seeds per specimen of \emph{A. perennans}, and $3.8$ seeds per specimen of \emph{E. virginicus}; we scored $10,342$ seeds in total. .
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each plant specimen.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive). 
$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})

@@

\subsection*{Modeling spatial and temporal changes in endophyte prevalence}
We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gaussian Models \cite{rue2009approximate}.
Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models.
We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equations (SPDE) to approximate a continuous spatial Gaussian process. 
This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.

@@

We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles encompassing the study area (Fig. \ref{fig:meshplot}).
Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
Priors for results presented in the main text reflect a range of $342$ kilometers. 
We found that model results were sensitive to this choice, and so tested a range of priors (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e. the same direction of effects across space, but that the magnitude and uncertainty varied. 
%In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.

\subsection*{Validating model performance with in-sample and out-of-sample tests}
We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and \tom{surveys of \emph{A. hyemalis} took place between 2015 and 2020}{We have added more recent AGHY survey data. I am not sure if you have access to this but you should definitely use it. Karl or I can point you to the right file.}.
Population surveys of \emph{A. hyemalis} were initially designed to cover longitudinal variation in endophyte prevalence towards its range edge, while surveys of \emph{E. virginicus} were designed to cover latitudinal variation along its range edge. 
In total, we visited 43 populations of \emph{A. hyemalis} and 20 populations of \emph{E. virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \ref{fig:contempsurveysmap}).
During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: $22.9$).
We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year. 
The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
We compared the model's prediction for these locations to the observed population prevalence.

@@

\subsection*{Modeling distributions of host species}
We modeled the distribution of each host species to generate maps on which we predicted the dynamics of \emph{Epichloë} symbionts.
We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}.
A full description of the ODMAP can be found in the  (Supplementary Method \ref{sec:sdm }). 
 In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale  of our selected climatic predictors. 
We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
%\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
The performance of models was evaluated with AUC \citep{jimenez2012insights}. 
We found AUC = 0.862, AUC = 0.838, AUC = 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}. 
These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence

@@

	\section*{Acknowledgments}
	We thank Jessica Budke for help in drafting our initial destructive sampling plan, and to the many members of herbarium staff who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections. 
	Several high schooler and undergraduate researchers contributed to data collection, including A. Appio-Riley, P. Bilderback, E. Chong, K. Dickens, L. Dufresne, B. Gutierrez, A. Johnson, S. Linder, E. Scales, B. Scherick, K. Schrader, E. Segal , G. Singla, and M. Tucker.
	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.

@@

	\section*{Statement of Authorship}
J.C.F. contributed to research conception, data collection, data analysis, and led manuscript drafting.
J.M. contributed to data analysis and manuscript manuscript revisions.
T.E.X.M. contributed to research conception, data collection, data analysis, and manuscript revisions.

	\section*{Data and Code Availability}
	Data from this publication will be made publically available at the time of publication and upon request.
	Code for analyses can be found through: https://github.com/joshuacfowler/EndoHerbarium

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = .8\linewidth]{density_plot.png}
	\caption{\textbf{Consistency between real data and simulated values indicate that the fitted model accurately describes the data}. Graph shows density curves for the observed data (red) along with the mean(solid) and 95\% CI (dashed) of simulated values (black).}
\end{figure}

@@

	\begin{table}[h]
		\caption{Summary of herbarium samples across collections}
		\label{Table:herbaria}
		\centering
		\begin{tabular}{llll}\hline
			Herbarium Collection        & AGHY        & AGPE      &      ELVI\\ \hline
			Botanical Research Institute of Texas &   341   &    189&    176    \\
			Louisiana State University &     71  & --  &   61       \\
			Mercer Botanic Garden &   3    & --     &     6\\
			Missouri Botanic Garden& 78 & 39& 31\\
		    Texas A\&M &  73&-- & 49 \\
		    University of Kansas & 134 & -- &  20\\
		    University of Oklahoma & 65 &30&  91\\
		    University of Texas  \& Lundell   &  169& 41& 99\\		    				 			     			     
			Oklahoma State University&     30  &   --    &  69 \\ \hline
		\end{tabular}
		\bigskip{}

	\end{table}	
	\section*{Supporting Methods}
	\subsection*{ODMAP Protocol}\label{sec:sdm}
{\color{blue}{Overview}}\\
\textbf {Model  purpose}: Mapping current distribution of epichloë host species. \\
\textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
\textbf {Study area}: Eastern North America \\
%Scale of analysis
\textbf {Spatial extent}: -125.0208, -66.47917, 24.0625, 49.9375 (xmin, xmax, ymin, ymax).\\
\textbf {Spatial resolution}: 0.04166667, 0.04166667 (x, y).\\
\textbf {Temporal extent}: 1990 to 2020.\\
\textbf {Boundary}: Natural.\\
{\color{blue}{Data}}\\
\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America. We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
\textbf{Response data type}: occurrence record, presence-only.\\
\textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
\textbf{Climatic data}:  raster data extracted from PRISM  \\
{\color{blue}{Model }}\\
\textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
\textbf{Algorithms}: Maximum entropy (maxent)\\
\textbf{Workflow}: We  described the workflow in the method section of the manuscript. \\
\textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
\textbf{Code availability}: Will be available upon acceptance \\
\textbf{Data availability}: Will be available upon acceptance \\
{\color{blue}{Assessment}}\\
We used AUC to test model performance.\\
{\color{blue}{Prediction }}\\
We predicted the probability of presence of the host species as a binary maps (presence or absence)

>>>>

	\textit{Manuscript elements}: Figure \ref{fig:map} - Figure \ref{fig:contemptestplot}, appendix~A (including Figure~\ref{fig:endo_status_map} - Figure \ref{fig:ELVI_climate_covariates},  Table \ref{table:herbaria}, and Supplemental Methods).

@@

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see Table \ref{table:herbaria} for a summary of specimens included from each collection). 
With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. \ref{fig:map}, Fig. \ref{fig:temporal}A, Fig. \ref{fig:endo_status_map}).
%Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centroid.

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
On average we scored $4.7$ seeds per specimen of \emph{A. hyemalis}, $4.2$ seeds per specimen of \emph{A. perennans}, and $3.8$ seeds per specimen of \emph{E. virginicus}; we scored $10,342$ seeds in total. .
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each plant specimen.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive). 
$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
The following analyses used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:FigA5})

@@

\subsection*{Modeling spatial and temporal changes in endophyte prevalence}
We assessed spatial and temporal changes in endophyte prevalence across each host distribution, quantifying the ``global'' temporal trends aggregated across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
To account for the spatial non-independence of geo-referenced occurrences, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
INLA provides a computationally efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gaussian Models \cite{rue2009approximate}.
Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models.
We incorporated spatial heterogeneity into this analysis using spatially-structured intercept and slope parameters implemented as stochastic partial differential equations (SPDE) to approximate a continuous spatial Gaussian process. 
This SPDE approach is a flexible method of smoothing across space while explicitly accounting for spatial dependence between data-points \citep{lindgren2011explicit,bakka2018spatial}.
Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies \cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.

@@

We performed model fitting using the inlabru R package \citep{bachl2019inlabru}.
Global intercept and slope parameters $\mathrm{A}$, and $\mathrm{T}$, were given vague priors.
Scorer and collector random effects, $\chi$ and $\omega$, were given penalized complexity priors, with distributions approximating a Normal distribution with standard deviation of 5. 
Each spatially-structured parameter depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles encompassing the study area (Fig. \ref{fig:meshplot}).
We assessed model fit with visual posterior predictive checks (\ref{fig:overlayplot}) and measurements of AUC (Figs. \ref{fig:ROCtraining}-\ref{fig:ROCtest}).
Priors, termed ''range" and "variance", define the distance of spatial decay described by the Mat\'{e}rn covariance function.
Priors for results presented in the main text reflect a range of $342$ kilometers. 
We found that model results were sensitive to this choice, and so tested a range of priors (from 68 kilometers to 2160 kilometers) and meshes (Supplemental Material), finding that model results were qualitatively similar, i.e. the same direction of effects across space, but that the magnitude and uncertainty varied. 
%In all cases, posterior modes were \tom{stable}{Assessed how?} and equal to zero, indicating successful numeric convergence.

\subsection*{Validating model performance with in-sample and out-of-sample tests}
We evaluated the predictive ability of the model using both in-sample training data from the herbarium surveys, and with out-of-sample test data from contemporary endophyte surveys, an important but rarely used strategy in ecological studies \cite{tredennick2021practical, lee2024phenological}.
We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} in Texas and the southern US. 
Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
%\tom{surveys of \emph{A. hyemalis} took place between 2015 and 2020}{We have added more recent AGHY survey data. I am not sure if you have access to this but you should definitely use it. Karl or I can point you to the right file.}.
Population surveys of \emph{A. hyemalis} were initially designed to cover longitudinal variation in endophyte prevalence towards its range edge, while surveys of \emph{E. virginicus} were designed to cover latitudinal variation along its range edge. 
In total, we visited 43 populations of \emph{A. hyemalis} and 20 populations of \emph{E. virginicus} across the south-central US, with emphasis on Texas and neighboring states (Fig \ref{fig:contempsurveysmap}).
During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: $22.9$).
We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys (with 5-10 seeds scored per individual), and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
For each population, we compared the observed fraction of endophyte-symbiotic hosts to the predicted probability of endophyte occurrence $\hat{P}$ derived from the model based on location and year. 
The contemporary survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
We compared the model's prediction for these locations to the observed population prevalence.

@@

\subsection*{Modeling distributions of host species}
We modeled the distribution of each host species to generate maps on which we predicted the dynamics of \emph{Epichloë} symbionts.
We followed the ODMAP (overview, data, model, assessment, prediction) protocol \citep{crossley2022opposing}.
A full description of the ODMAP can be found in the  (Supporting Methods). 
 In short, we used presence-only observations of the host species from Global Biodiversity Information Facility (GBIF) between 1990 to 2020.
To reduce the potential influence of sampling bias and spatial autocorrelation, we thinned the occurrences to the spatial scale  of our selected climatic predictors. 
We selected climate variables that aligned with our analysis of climatic influences on trends in endophyte prevalence described above.
We calculated the mean and standard deviation of seasonal temperature and precipitation across 1990 to 2020. 
Among this suite of variables, we chose to include, which were uncorrelated (Variance Inflation Factor $>$ 0.7) and allowed us to predict the occurrence probability of each host species in space and time.
These climatic variables are: the mean and standard deviation of spring, summer, and autumn temperature, as well as the mean and standard deviation of spring, summer, and autumn cumulative precipitation, cumulative precipitation, and cumulative precipitation.
We fit maximum entropy (MaxEnt) models using the maxent function in the package dismo \citep{hijmans2017package}. 
%\josh{MaxEnt is preferred because it has been shown to generate response curves with less unpredictable behavior when applied to new climates \citep{hijmans2006ability}.}{possibly could remove this sentence?}
We generated 10,000 pseudo-absences as background points, and split the occurrence data into 75\% for model training and 25\% for model testing.
The performance of models was evaluated with AUC \citep{jimenez2012insights}. 
We found AUC = 0.862, AUC = 0.838, AUC = 0.821 respectively  for \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
To convert the continuous predicted probabilities into binary presence - absence maps, we used the  training sensitivity (true positive rate) and specificity threshold (true negative rate) \citep{liu2005selecting}. 
These binary maps serve as boundaries in presented maps of change in endophyte prevalence, and outline the set of pixels used in our analysis of climate correlates with trends in endophyte prevalence

@@

	\section*{Acknowledgments}
	We thank Dr. Jessica Budke for help in drafting our initial destructive sampling plan, and to the many staff members of herbaria who facilitated our research visits, as well as to the hundreds of collectors who contributed to the natural history collections. 
	Several high schooler and undergraduate researchers contributed to data collection, including A. Appio-Riley, P. Bilderback, E. Chong, K. Dickens, L. Dufresne, B. Gutierrez, A. Johnson, S. Linder, E. Scales, B. Scherick, K. Schrader, E. Segal , G. Singla, and M. Tucker.
	This research was supported by funding from National Science Foundation (grants 1754468 and 2208857) and by funding from the Texas Ecolab Program.

@@

	\section*{Statement of Authorship}
J.C.F. contributed to research conception, data collection, data analysis, and led manuscript drafting.
J.M. contributed to data analysis and manuscript revisions.
T.E.X.M. contributed to research conception, data collection, data analysis, and manuscript revisions.

	\section*{Data and Code Availability}
	Data from this publication will be made publically available upon acceptance and before that upon request.
	Code for analyses can be found through: https://github.com/joshuacfowler/EndoHerbarium

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = .5\linewidth]{../Plots/overlay_plot.png}
	\caption{\textbf{Consistency between real data and simulated values indicate that the fitted model accurately describes the data}. Graph shows density curves for the observed data (black) along with with 100 simulated datasets (red).}
			\label{fig:overlayplot}
\end{figure}

@@

	\begin{table}[h]
		\caption{Summary of herbarium samples across collections}
		\label{table:herbaria}
		\centering
		\begin{tabular}{llll}\hline
			Herbarium Collection        & AGHY        & AGPE      &      ELVI\\ \hline
			Botanical Research Institute of Texas &   350   &    190&    198    \\
			Louisiana State University &     72  & 38  &   62       \\
			Mercer Botanic Garden &   3    & --     &     6\\
			Missouri Botanic Garden& 210 & 205 & 122\\
		    Texas A\&M &  100 &-- & 72 \\
		    University of Kansas & 134 & 34 &  197\\
		    University of Oklahoma & 85 &34&  95\\
		    University of Texas  \& Lundell   &  183& 91& 102\\		    				 			     			     
			Oklahoma State University&     51  &   10    &  74 \\ \hline
		\end{tabular}
		\bigskip{}

	\end{table}	
	\section*{Supporting Methods}
	\subsection*{ODMAP Protocol}\label{sec:sdm}
{\color{blue}{Overview}}\\
\textbf {Model  purpose}: Mapping current distribution of \emph{Epichloë} host species. \\
\textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
\textbf {Study area}: Eastern North America \\
%Scale of analysis
\textbf {Spatial extent}: -125.0208, -66.47917, 24.0625, 49.9375 (xmin, xmax, ymin, ymax).\\
\textbf {Spatial resolution}: 0.04166667, 0.04166667 (x, y).\\
\textbf {Temporal extent}: 1990 to 2020.\\
\textbf {Boundary}: Natural.\\
{\color{blue}{Data}}\\
\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America. We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
\textbf{Response data type}: occurrence record, presence-only.\\
\textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
\textbf{Climatic data}:  raster data extracted from PRISM  \\
{\color{blue}{Model }}\\
\textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
\textbf{Algorithms}: Maximum entropy (maxent)\\
\textbf{Workflow}: We  described the workflow in the method section of the manuscript. \\
\textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium \\
\textbf{Data availability}: Will be available upon acceptance \\
{\color{blue}{Assessment}}\\
We used AUC to test model performance.\\
{\color{blue}{Prediction }}\\
We predicted the probability of presence of the host species as a binary maps (presence or absence)
